Double Standards in Judging Collective Action

### 2022-09-13 Documents/manuscript.tex
preprint

\documentclass[twocolumn, 11pt, letterpaper]{article}

% Font
\usepackage[utf8]{inputenc}
\usepackage{MinionPro} 
\input{glyphtounicode}
\pdfgentounicode=1 
\usepackage{microtype}

% Format
\usepackage[letterpaper, margin = 1in]{geometry}
\setlength{\columnsep}{18pt}
\setcounter{secnumdepth}{0}
\usepackage[compact]{titlesec}
\titleformat*{\section}{\centering\normalfont\Large\bfseries}
\titleformat*{\subsection}{\raggedright\normalfont\normalsize\bfseries}
\titleformat*{\subsubsection}{\raggedright\normalfont\normalsize\bfseries\itshape}
\usepackage{authblk}
\renewcommand\Affilfont{\small}
\newcommand{\refsection}{\section{References}}
\providecommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

@@

% Frontmatter
\title{ Double Standards in Judging Collective Action\thanks{} }
\author[1]{Nils K. Reimer}
\affil[1]{University of California, Santa Barbara, United States}
\author[2]{Marija Brankovi\'{c}}
\affil[2]{Singidunum University, Serbia}
\author[3]{Iniobong Essien}
\affil[3]{Leuphana University of L\"{u}neburg, Germany}
\author[4]{Jin X. Goh}
\affil[4]{Colby College, United States}
\author[5]{S\'{e}bastien Goudeau}
\affil[5]{University of Poitiers, France}
\author[6]{N\'{o}ra A. Lantos}
\affil[6]{E\"{o}tv\"{o}s Lor\'{o}nd University, Hungary}
\author[7]{Jenny Veldman}
\affil[7]{KU Leuven, Belgium}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      \noindent Collective action is a powerful force driving social
change but often sparks contention about what actions are acceptable
means to effect social change. In five studies (total \emph{N} = 2,979),
we investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no consistent ingroup bias when judging collective action by
ingroup and outgroup members. Our findings have theoretical and
practical implications for understanding the often divided response to
prominent social movements.
    \end{abstract}
  \end{@twocolumnfalse}
  % \renewcommand{\abstractname}{Statement of Relevance}
  % \begin{@twocolumnfalse}
  %   \maketitle
  %   \begin{abstract}
  %   \noindent 
  %   \end{abstract}
  % \end{@twocolumnfalse}
]

{
  \renewcommand{\thefootnote}%
    {\fnsymbol{footnote}}
  \footnotetext[1]{M.B.--J.V. are listed in alphabetical order. This research was supported by a Seedcorn Research Grant from the European Association of Social Psychology.}
}

\noindent Protest movements often spark contention about what actions
are acceptable means of protest. Reactions to Black American athletes
kneeling during the national anthem to protest racist police violence
are a case in point. Kneeling during the national anthem is not violent,
disruptive, or illegal. And yet, only 29\% of White Americans (66\% of
Black Americans) and 11\% of Republicans (59\% of Democrats) considered
it appropriate for Black athletes to protest in this way (YouGov, 2017).
Indeed, Black athletes who had participated in the protests faced a
backlash in public opinion (Lacina, 2020). This and other examples
suggest that \emph{who} the protesters are and \emph{what} they are
protesting influences how acceptable observers judge their actions to
be.

@@

Collective action can take many forms. Psychologists distinguish between
\emph{normative} collective action that seeks to achieve a political
goal while conforming to the norms of the existing social system and
\emph{non-normative} collective action that violates those norms (Wright
et al., 1990). Recent research has shown that this distinction matters
for how people respond to collective action. Shuman et al. (2021) showed
that non-normative, non-violent collective action is more effective than
either normative or violent collective action at gaining concessions
from those most resistant to social change. Feinberg et al. (2020)
demonstrated, however, that observers are less supportive of social
movements that use extreme protest actions. Similarly, Teixeira et al.
(2020) found that advantaged-group members perceive non-normative
collective action by disadvantaged-group members as more damaging to
their ingroup's social image and are thus less supportive of such
action. Verkuyten et al. (2022) found that, in general, the public tends
to be intolerant of non-normative collective action. This means that
activists face the dilemma that non-normative collective action might be
most effective at gaining concessions but that it also risks reducing
popular support for a cause. Together, these studies show that the
distinction between normative and non-normative collective action is
psychologically and societally consequential.

Research has, for the most part, relied on ad-hoc distinctions between
what researchers themselves considered normative and non-normative
collective action. In liberal democracies, researchers tend to consider
actions such as signing petitions, voting in elections, or peaceful
protest to be normative and actions such as blocking traffic, damaging
property, or violent protest to be non-normative. But, as the examples
in the first paragraph show, the distinction between normative and
non-normative collective action might instead be in the eye of the
beholder.

In this research, we move this distinction from the researcher's
intuition into the realm of scientific investigation and test for double
standards in where people draw the line between acceptable and
unacceptable forms of collective action. For that purpose, we define a
double standard as judging the same action as more or less acceptable
depending on who is performing the action and for what reason (for an
analogous definition, see Foschi, 2000).\footnote{Feinberg et al. (2020)
  provided incidental evidence for double standards in judging
  collective action as, in a manipulation check, Black participants
  rated Black Lives Matter protests as less extreme than White
  participants and liberal participants rated protest actions for
  progressive causes as less extreme than conservative participants.} In
this section, we derive predictions from social psychological theories
about how identity-based and ideology-based double standards could lead
people to judge the same protest actions as more or less acceptable
depending on who the protesters are and what they are protesting.

@@

One domain in which ingroup bias manifests is judgements about other
people's actions. Hewstone (1990) reviewed research showing that people
make ingroup-serving causal attributions when judging actions by ingroup
and outgroup members, for example, by attributing negative behavior by
outgroup members to internal causes but attributing negative behavior by
ingroup members to external causes. Valdesolo \& DeSteno (2007)
demonstrated that participants judged the same selfish action as less
unfair when performed by themselves or an ingroup member than when
performed by an outgroup member. Other studies (Abrams et al., 2013;
Endevelt et al., 2021; for exceptions, see Mendoza et al., 2014; Pinto
et al., 2010) provided further evidence for ingroup bias in judging
moral transgressions by ingroup and outgroup members.

@@

System Justification Theory (Jost \& Banaji, 1994; for a recent review,
see Jost, 2020) argues that people are motivated to defend, justify, and
bolster the prevailing social, economic, and political system because
doing so serves basic epistemic, existential, and relational needs. Both
advantaged and disadvantaged groups are thought to be motivated to
justify the existing system, although the strength of this motivation
and its expression vary across individuals and situations.

@@

Just as system justification shapes support for, and opposition to,
collective action, we propose that it results in ideology-based double
standards in what actions people consider acceptable forms of collective
action. First, we hypothesize that people will judge the same protest
actions as more acceptable if the protest supports, rather than
challenges, the system (\emph{Hypothesis 2a}). This hypothesis reflects
the assumption that the motivation to justify the system is, to some
extent, universal. Second, we hypothesize that people will judge the
same protest actions as more acceptable if they endorse
system-justifying beliefs and the protest supports the system or if they
reject system-justifying beliefs and the protest challenges the system
(\emph{Hypothesis 2b}). Notably, this hypothesis implies a symmetry in
how people with different ideological orientations judge
system-supporting and system-challenging collective action.

@@

Item response theory is a a conceptual and statistical framework for
understanding how the characteristics of both items and respondents
shape responses to a set of items (DeMars, 2010). When respondents judge
whether each of a set of actions is an acceptable means of protest (1 =
\emph{yes}, 0 = \emph{no}), a two parameter logistic item response
theory model estimates responses as a function of several latent
(unobserved) parameters:
\[ \Pr ( y_{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) \]
where \(\Pr ( y_{ij} = 1 )\) is the probability that respondent \(j\)
considers action \(i\) an acceptable means of protest, \(\theta_j\)
estimates how \emph{accepting} respondent \(j\) is of various protest
actions, \(\beta_i\) estimates how \emph{acceptable} action \(i\) is,
and \(\alpha_i\) estimates how \emph{discriminating} action \(i\) is.

@@

In this research, item response theory serves two purposes. First, item
response theory provides a statistical framework for scale development.
In Studies 1 and 2, we select the most discriminating and informative
actions to build an instrument for capturing double standards in the
experiments. Second, item response theory provides a formal and
statistical definition of double standards in judging collective action
as respondents being more accepting of the same protest actions
depending on who the protesters and what they are protesting. In
Experiments 1--3, we operationalize double standards as the differences
in \(\theta_{(z)}\) between conditions which, as \(\theta_{(z)}\) is
\(z\)-standardized, correspond to Cohen's \(d\) effect sizes.

@@

The first purpose of our research is to develop an instrument to measure
where people draw the line between acceptable and unacceptable forms of
collective action. In Study 1, we ask participants to generate more and
less extreme protest actions and compile a pool of protest actions for
scale development. In Study 2, we ask participants to rate how
acceptable they consider each of those protest action to be and use item
response theory to select the most discriminating and informative
protest actions for our instrument. In so doing, we build an instrument
that enables both present and future research to capture double
standards in judging collective action.

@@

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that, in line with \emph{Hypothesis
1a} and \emph{1b}, participants would judge the same protest actions as
more acceptable if the protesters were ingroup members protesting for
the ingroup's rights than if they were outgroup members protesting for
the outgroup's rights. By comparing reactions to working-class and
middle-class protesters, we tested how observers judge collective action
by lower-status (i.e.~working-class) and higher-status (middle-class)
group members. Assuming that collective action by lower-status group
members is seen as more threatening to the unequal status quo than
collective action by higher-status group members and that people are, to
same extent, motivated to justify the system, we tested the
preregistered hypothesis that, in line with \emph{Hypothesis 2a},
participants would judge the same protest actions as more acceptable if
the protesters were from a higher-status group than if they were from a
lower-status group. In non-preregistered analyses, we further explored
whether, in line with \emph{Hypothesis 2b}, participants would judge the
same protest actions as more or less acceptable depending on their own
ideological positions. In this way, Experiment 1 provided a first test
of the hypothesized identity- and ideology-based double standards in
judging collective action.

@@

We recruited 515 participants from the Prolific subject pool who were UK
citizens, 25 years old or older, and not current students.\footnote{We
  had preregistered a sample of 500 participants, before exclusions, but
  included another 15 participants who completed the study without
  returning an approval code to the Prolific platform. Participants
  received, on average, \pounds 9.00 (\$10.93) per hour of
  participation.} As preregistered, we excluded 71 participants who
failed an attention check. This resulted in a final sample of 443
participants (\(\textit{Mdn} = 41\) years, age range: 25--76 years; 272
women, 171 men, 1 non-binary). Of these, 210 participants considered
their past, current, or future jobs to be working-class jobs.
Participants in this lower-status group did not have a university degree
and placed themselves on the bottom three ranks of the subjective
socioeconomic status ladder. Another 233 participants considered their
past, current job, or future jobs to be middle-class/professional jobs.
Participants in this higher-status group had at least an undergraduate
degree and placed themselves on the top four ranks of the subjective
socioeconomic status ladder.

@@

We assessed reactions to the vignette by asking participants how
outraged they would be if the government were to introduce this bill in
real life, to what extent it would affect them personally, and to what
extent the it would affect people like them (1 = \emph{not at all}, 7 =
\emph{very much}). We also asked participants to what extent they
identified with people with the kinds of jobs most affected by the
proposed bill (1 = \emph{not at all}, 7 = \emph{very much}) and how
often, if at all, they had participated in protest actions such as the
ones we asked about (1 = \emph{never}, 5 = \emph{very often}).

@@

Figure \ref{fig:f3} (B) the \(z\)-standardized propensity to consider
more controversial protest actions acceptable as a function of the three
ideological orientation variables. We found that participants who
reported a more right-wing political orientation
(\(\beta_{xy} = -0.25, [-0.37, -0.15]\)) and who expressed more
agreement with system-justifying beliefs
(\(\beta_{xy} = -0.20, [-0.30, -0.10]\)) tended to find fewer collective
actions to be acceptable. In contrast, we found that, after controlling
for the other two variables, social dominance orientation was not
associated with participants' judgements about how acceptable various
collective actions are (\(\beta_{xy} = -0.05, [-0.15, 0.05]\)). Overall,
these findings suggest that people who are right-wing and endorse
system-justifying beliefs tend to find various collective actions to be
less acceptable than people who are left-wing and do not endorse
system-justifying beliefs.\footnote{We ran additional analyses to
  explore how these associations differed across experimental
  conditions. We found uncertainty intervals for these associations to
  overlap across conditions, suggesting that we did not have enough data
  to differentiate these varying effects.}

@@

In Experiment 2, we investigated potential double standards in judging
collective action for or against defunding the police in the United
States. That is, we tested whether Black and White Americans applied
different standards when judging collective action taken by either Black
or White protesters to protest either for or against police divestment
as a possible solution to racist police violence. We conducted this
study in January 2021, a time when, after a year of unprecedented
protests for racial justice, most Americans could be expected to be
aware of, and have formed an opinion on, the Black Lives Matter
movement. We focused on defunding the police as this position remained
controversial among both Black and White Americans even as they broadly
agreed on the need for police reform.\footnote{In December 2020, only
  34\% of Black Americans and 18\% of White Americans supported
  defunding the police while majorities of Black and White Americans
  supported police reforms such as eliminating qualified immunity (50\%
  and 57\%) or banning chokeholds (79\% and 61\%, YouGov, 2020).}

@@

As expected, Black participants (\(M = 6.18, \textit{SD} = 1.31\))
reported being more outraged about recent incidents of police violence
than White participants (\(M = 5.56, \textit{SD} = 1.61\); Cohen's
\(d = 0.42\)). Black participants strongly identified with their ingroup
(\(M = 6.46, \textit{SD} = 1.09\)) and, across experimental conditions,
identified more with Black protesters (\(M = 4.66, \textit{SD} = 1.99\))
than with White protesters (\(M = 4.03, \textit{SD} = 2.16\);
\(d = 0.30\)). In contrast, White participants identified less strongly
with their ingroup (\(M = 4.85, \textit{SD} = 1.53\); \(d = 1.21\)) and
did not identify more with White protesters
(\(M = 3.58, \textit{SD} = 2.01\)) than with Black protesters
(\(M = 3.81, \textit{SD} = 1.77\); \(d = -0.12\)). Both White
(\(d = 0.67\)) and Black (\(d = 0.53\)) participants tended to identify
more with protesters protesting \emph{for} defunding the police. On
average, participants tended to describe their political orientation as
moderately liberal (\(M = 2.92, \textit{SD} = 1.73\)) but, as expected,
were divided in their support for defunding the police
(\(M = 3.37, \textit{SD} = 1.42\)). That is, 56\% supported defunding
the police, 30\% opposed defunding the police, and 14\% remained
undecided.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a
function of their \emph{z}-standardized endorsement of system-justifying
beliefs. As preregistered, we modeled this relationship with two fixed
effects, one estimating the effect of system-justifying beliefs on
judgements about system-defending protest actions and one estimating
their effect on judgements about system-challenging protest actions, and
one varying effect estimating its variance across conditions. Supporting
\emph{Hypothesis 2b}, participants who \emph{rejected} system-justifying
beliefs were \emph{more} likely to consider system-challenging protest
actions (for defunding the police) to be acceptable means of protest
(\(\beta_{xy} = 0.38, [0.16, 0.58]\)). We did not, however, find
evidence for the ideological symmetry implied by \emph{Hypothesis 2b}
since participants who \emph{endorsed} system-justifying beliefs were
\emph{not} more likely to consider system-defending protest actions
(against defunding the police) to be acceptable means of protest
(\(\beta_{xy} = -0.18, [-0.40, 0.02]\)).

@@

We explored whether group identification moderated how the participants'
and the protesters' group memberships affected participants' responses.
To that end, we extended Model 1 by estimating participants' responses
as a function of their \emph{z}-standardized identification with their
racial ingroup. As in Model 2, we modeled this relationship with a fixed
and a varying effect. We found, however, that even participants who
strongly identified with their racial ingroup (+1\emph{SD}) did not
consider protest actions performed by their ingroup to be more
acceptable than the same actions performed by the relevant outgroup
(\(d = 0.01, [-0.00, 0.02]\)) and did not consider protest actions for a
cause that was aligned with their ingroup's interests to be more
acceptable than protest actions for a cause not aligned with their
ingroup's interests (\(d = 0.00, [-0.01, 0.01]\)). Our non-preregistered
analyses thus suggested that group identification did not moderate group
differences in judgements about collective action by ingroup and
outgroup members.

@@

In Experiment 3, we investigated potential double standards in judging
collective action for or against restricting abortion in the United
States. That is, we tested whether liberal and conservative women and
men applied different standards when judging collective action opposing
or supporting further restricting or banning abortion. We conducted this
study in June 2022 when many expected the Supreme Court to overturn its
\emph{Roe v. Wade} decision that hitherto had guaranteed access to legal
abortions in the first 20--24 weeks of pregnancy in the United States.
We focused on restricting abortion as it is a conservative issue, as it
primarily affects women's reproductive rights, and as the anticipated
Supreme Court decision would prompt several states to restrict or ban
abortion.

>>>>

\documentclass[12pt, letterpaper]{article}

% Font
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{microtype}

% Format
\usepackage[letterpaper, margin = 1in]{geometry}
\setcounter{secnumdepth}{0}
\setlength{\parindent}{0.5in}
\usepackage[compact]{titlesec}
\titleformat*{\section}{\centering\normalfont\normalsize\bfseries}
\titleformat*{\subsection}{\raggedright\normalfont\normalsize\bfseries}
\titleformat*{\subsubsection}{\raggedright\normalfont\normalsize\bfseries\itshape}
\newcommand{\refsection}{\newpage \section{References}}
\usepackage{indentfirst}
\usepackage[hang,flushmargin]{footmisc}
\providecommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Header
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{14.5pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{DOUBLE STANDARDS}
\fancyhead[R]{\thepage}

@@

% Line spacing
\usepackage[doublespacing]{setspace}

% Frontmatter
\title{ Double Standards in Judging Collective Action }

@@

\section{Abstract}

\noindent Collective action is a powerful force driving social change
but often sparks contention about what actions are acceptable means to
effect social change. In five studies (total \emph{N} = 2,979), we
investigated double standards in judging collective action---that is,
whether observers will judge the same protest actions as more acceptable
if the protesters are ingroup members and their cause aligns with the
ingroup's interests (\emph{identity-based double standards}) or if the
protesters' cause aligns with the observers' ideological positions
(\emph{ideology-based double standards}). In two studies, we developed
an instrument of 25 controversial protest actions, based in item
response theory, to measure where people draw the line between
acceptable and unacceptable forms of collective action. In three
preregistered experiments, we used this instrument to test for double
standards in judging collective action for workers' rights in the United
Kingdom and for or against defunding the police and for or against
restricting abortion in the United States. We found evidence for
ideology-based, but not for identity-based, double standards:
Participants judged the same protest actions to be more acceptable when
the protesters' cause aligned with their own ideological position---but
showed no consistent ingroup bias when judging collective action by
ingroup and outgroup members. Our findings have theoretical and
practical implications for understanding the often divided response to
prominent social movements.

\textit{Keywords:} protest, collective action, double standards, social
identity theory, system justification theory, Black Lives Matter

\section{Double Standards in Judging Collective Action}

Protest movements often spark contention about what actions are
acceptable means of protest. Reactions to Black American athletes
kneeling during the national anthem to protest racist police violence
are a case in point. Kneeling during the national anthem is not violent,
disruptive, or illegal. And yet, only 29\% of White Americans (compared
to 66\% of Black Americans) and 11\% of Republicans (compared to 59\% of
Democrats) considered it appropriate for Black athletes to protest in
this way (YouGov, 2017). Indeed, Black athletes who had participated in
the protests faced a backlash in public opinion (Lacina, 2020). This and
other examples suggest that \emph{who} the protesters are and
\emph{what} they are protesting influences how acceptable observers
judge their actions to be.

@@

Collective action can take many forms. Psychologists distinguish between
\emph{normative} collective action that seeks to achieve a political
goal while conforming to the norms of the existing social system and
\emph{non-normative} collective action that violates those norms (Becker
\& Tausch, 2015; Wright et al., 1990). Some instead refer to the two
kinds of collective action as moderate and radical collective action
(e.g., Jiménez-Moya et al., 2015). Broadly, this distinction captures
the intuition that there are some forms of collective action that are
generally considered acceptable and others that are not.

Recent research has shown that this distinction matters for how people
respond to collective action. Shuman et al. (2021) showed that
non-normative, non-violent collective action is more effective than
either normative or violent collective action at gaining concessions
from those most resistant to social change. Feinberg et al. (2020)
demonstrated, however, that observers are less supportive of social
movements that use extreme protest actions. Similarly, Teixeira et al.
(2020) found that advantaged-group members perceive non-normative
collective action by disadvantaged-group members as more damaging to
their ingroup's social image and are thus less supportive of such
action. Verkuyten et al. (2022) found that, in general, the public tends
to be intolerant of non-normative collective action. This means that
activists face the dilemma that non-normative collective action might be
most effective at gaining concessions but that it also risks reducing
popular support for a cause. Together, these studies show that the
distinction between normative and non-normative collective action is
psychologically and societally consequential.

These and other studies have, for the most part, relied on ad-hoc
distinctions between what researchers themselves considered normative
and non-normative collective action. In liberal democracies, researchers
tend to consider actions such as signing petitions, voting in elections,
or peaceful protest to be normative and actions such as blocking
traffic, damaging property, or violent protest to be non-normative. But,
as reactions to Black athletes kneeling in protest suggest, the same
action might be considered acceptable by some and unacceptable by
others. That is, the distinction between normative and non-normative
collective action might be in the eye of the beholder.

In this research, we move this distinction from the researcher's
intuition into the realm of scientific investigation and test for double
standards in where people draw the line between acceptable (normative)
and unacceptable (non-normative) forms of collective action. For that
purpose, we define a double standard as judging the same action as more
or less acceptable depending on who is performing the action and for
what reason (for an analogous definition, see Foschi, 2000).\footnote{Feinberg
  et al. (2020) provided incidental evidence for double standards in
  judging collective action as, in a manipulation check, Black
  participants rated Black Lives Matter protests as less extreme than
  White participants and liberal participants rated protest actions for
  progressive causes as less extreme than conservative participants.} In
this section, we derive predictions from social psychological theories
about how identity-based and ideology-based double standards could lead
people to judge the same protest actions as more or less acceptable
depending on who the protesters are and what they are protesting.

@@

One domain in which ingroup bias manifests is judgments about other
people's actions. Hewstone (1990) reviewed research showing that people
make ingroup-serving causal attributions when judging actions by ingroup
and outgroup members, for example, by attributing negative behavior by
outgroup members to internal causes but attributing negative behavior by
ingroup members to external causes (see Pettigrew, 1979). Valdesolo \&
DeSteno (2007) demonstrated that participants judged the same selfish
action as less unfair when performed by themselves or an ingroup member
than when performed by an outgroup member. Other studies (Abrams et al.,
2013; Endevelt et al., 2021; for exceptions, see Mendoza et al., 2014;
Pinto et al., 2010) provided further evidence for ingroup bias in
judging moral transgressions by ingroup and outgroup members.

@@

System Justification Theory (Jost et al., 2004; for a recent review, see
Jost, 2020) argues that people are motivated to defend, justify, and
bolster the prevailing social, economic, and political system because
doing so serves basic epistemic, existential, and relational needs. Both
advantaged and disadvantaged groups are thought to be motivated to
justify the existing system, although the strength of this motivation
and its expression vary across individuals and situations.

@@

Just as system justification affects support for, and opposition to,
collective action, we propose that it results in ideology-based double
standards in what actions people consider acceptable forms of collective
action. First, we hypothesize that people will judge the same protest
actions as more acceptable if the protest supports, rather than
challenges, the system (\emph{Hypothesis 2a}). This hypothesis reflects
the assumption that the motivation to justify the system is, to some
extent, universal. Second, we hypothesize that people will judge the
same protest actions as more acceptable if they endorse
system-justifying beliefs and the protest supports the system or if they
reject system-justifying beliefs and the protest challenges the system
(\emph{Hypothesis 2b}). Notably, this hypothesis implies a symmetry in
how people with different ideological orientations judge
system-supporting and system-challenging collective action.

@@

Item response theory is a conceptual and statistical framework for
understanding how the characteristics of both items and respondents
shape responses to a set of items (DeMars, 2010). When respondents judge
whether each of a set of actions is an acceptable means of protest (1 =
\emph{yes}, 0 = \emph{no}), a two-parameter logistic item response
theory model estimates responses as a function of several latent
(unobserved) parameters:
\[ \Pr ( y_{ij} = 1 ) = \text{logit}^{-1} ( \alpha_i ( \theta_j + \beta_i ) ) \]
where \(\Pr ( y_{ij} = 1 )\) is the probability that respondent \(j\)
considers action \(i\) an acceptable means of protest, \(\theta_j\)
estimates how \emph{accepting} respondent \(j\) is of various protest
actions, \(\beta_i\) estimates how \emph{acceptable} action \(i\) is,
and \(\alpha_i\) estimates how \emph{discriminating} action \(i\) is.

@@

In this research, item response theory serves two purposes. First, item
response theory provides a statistical framework for scale development.
In Studies 1 and 2, we select the most discriminating and informative
actions to build an instrument for capturing double standards in judging
collective action. Second, item response theory provides a formal and
statistical definition of double standards in judging collective action
as respondents being more accepting of the same protest actions
depending on who the protesters are and what they are protesting. In
Experiments 1--3, we operationalize double standards as the differences
in \(\theta_{(z)}\) between conditions which, as \(\theta_{(z)}\) is
\(z\)-standardized, correspond to Cohen's \(d\) effect sizes.

@@

The first purpose of our research is to develop an instrument to measure
where people draw the line between acceptable and unacceptable forms of
collective action. In Study 1, we ask participants to generate more and
less extreme protest actions and compile a pool of protest actions for
scale development. In Study 2, we ask participants to rate how
acceptable they consider each of those protest actions to be and use
item response theory to select the most discriminating and informative
protest actions for our instrument. In so doing, we build an instrument
that enables both present and future research to capture double
standards in judging collective action.

@@

By varying the participants' and the protesters' group memberships, we
tested the preregistered hypothesis that, in line with \emph{Hypothesis
1a} and \emph{1b}, participants would judge the same protest actions as
more acceptable if the protesters were ingroup members protesting for
the ingroup's rights than if they were outgroup members protesting for
the outgroup's rights. By comparing reactions to working-class and
middle-class protesters, we tested how observers judge collective action
by lower-status (i.e.~working-class) and higher-status (middle-class)
group members. Assuming that collective action by lower-status group
members is seen as more threatening to the unequal status quo than
collective action by higher-status group members and that people are, to
same extent, motivated to justify the system, we tested the
preregistered hypothesis that, in line with \emph{Hypothesis 2a},
participants would judge the same protest actions as more acceptable if
the protesters were from a higher-status group than if they were from a
lower-status group. In non-preregistered analyses, we further explored
whether participants would judge the same protest actions as more or
less acceptable depending on their own ideological positions. In this
way, Experiment 1 provided a first test of the hypothesized identity-
and ideology-based double standards in judging collective action.

@@

We recruited 515 participants from the Prolific subject pool who were UK
citizens, 25 years old or older, and not current students.\footnote{We
  had preregistered a sample of 500 participants, before exclusions, but
  included another 15 participants who completed the study without
  returning an approval code to the Prolific platform. Participants
  received, on average, \pounds 9.00 (\$10.93) per hour of
  participation.} As preregistered, we excluded 71 participants who
failed an attention check. This resulted in a final sample of 443
participants (\(\textit{Mdn} = 41\) years, age range: 25--76 years; 272
women, 171 men, 1 non-binary). Of these, 210 participants considered
their past, current, or future jobs to be working-class jobs.
Participants in this lower-status group did not have a university degree
and placed themselves on the bottom three ranks of the subjective
socioeconomic status ladder. Another 233 participants considered their
past, current, or future jobs to be middle-class/professional jobs.
Participants in this higher-status group had at least an undergraduate
degree and placed themselves on the top four ranks of the subjective
socioeconomic status ladder.

@@

We assessed reactions to the vignette by asking participants how
outraged they would be if the government were to introduce this bill in
real life, to what extent it would affect them personally, and to what
extent it would affect people like them (1 = \emph{not at all}, 7 =
\emph{very much}). We also asked participants to what extent they
identified with people with the kinds of jobs most affected by the
proposed bill (1 = \emph{not at all}, 7 = \emph{very much}) and how
often, if at all, they had participated in protest actions such as the
ones we asked about (1 = \emph{never}, 5 = \emph{very often}).

@@

Figure \ref{fig:f3} (B) the \(z\)-standardized propensity to consider
more controversial protest actions acceptable as a function of the three
ideological orientation variables. We found that participants who
reported a more right-wing political orientation
(\(\beta_{xy} = -0.25, [-0.37, -0.15]\)) and who expressed more
agreement with system-justifying beliefs
(\(\beta_{xy} = -0.20, [-0.30, -0.10]\)) tended to find fewer collective
actions to be acceptable. In contrast, we found that, after controlling
for the other two variables, social dominance orientation was not
associated with participants' judgments about how acceptable various
collective actions are (\(\beta_{xy} = -0.05, [-0.15, 0.05]\)). Overall,
these findings suggest that people who are right-wing and endorse
system-justifying beliefs tend to find various collective actions to be
less acceptable than people who are left-wing and do not endorse
system-justifying beliefs.\footnote{We ran additional analyses to
  explore how these associations differed across experimental
  conditions. We found uncertainty intervals for these associations to
  overlap across conditions, suggesting that we did not have enough data
  to differentiate these varying effects.}

@@

In Experiment 2, we investigated potential double standards in judging
collective action for or against defunding the police in the United
States. That is, we tested whether Black and White Americans applied
different standards when judging collective action taken by either Black
or White protesters to protest either for or against police divestment
as a possible solution to racist police violence. We conducted this
study in January 2021, a time when, after months of unprecedented
protests for racial justice, most Americans could be expected to be
aware of, and have formed an opinion on, the Black Lives Matter
movement. We focused on defunding the police as this position remained
controversial among both Black and White Americans even as they broadly
agreed on the need for police reform.\footnote{In December 2020, only
  34\% of Black Americans and 18\% of White Americans supported
  defunding the police while majorities of Black and White Americans
  supported police reforms such as eliminating qualified immunity (50\%
  and 57\%) or banning chokeholds (79\% and 61\%, YouGov, 2020).}

@@

As expected, Black participants (\(M = 6.18, \textit{SD} = 1.31\))
reported being more outraged about recent incidents of police violence
than White participants (\(M = 5.56, \textit{SD} = 1.61\); Cohen's
\(d = 0.42\)). Black participants strongly identified with their ingroup
(\(M = 6.46, \textit{SD} = 1.09\)) and, across experimental conditions,
identified more with Black protesters (\(M = 4.66, \textit{SD} = 1.99\))
than with White protesters (\(M = 4.03, \textit{SD} = 2.16\);
\(d = 0.30\)). In contrast, White participants identified less strongly
with their ingroup than Black participants
(\(M = 4.85, \textit{SD} = 1.53\); \(d = 1.21\)) and did not identify
more with White protesters (\(M = 3.58, \textit{SD} = 2.01\)) than with
Black protesters (\(M = 3.81, \textit{SD} = 1.77\); \(d = -0.12\)). Both
White (\(d = 0.67\)) and Black (\(d = 0.53\)) participants tended to
identify more with protesters protesting \emph{for} defunding the
police. On average, participants tended to describe their political
orientation as moderately liberal (\(M = 2.92, \textit{SD} = 1.73\))
but, as expected, were divided in their support for defunding the police
(\(M = 3.37, \textit{SD} = 1.42\)). That is, 56\% supported defunding
the police, 30\% opposed defunding the police, and 14\% remained
undecided.

@@

Model 2 (M2) extended Model 1 by estimating participants' responses as a
function of their \emph{z}-standardized endorsement of system-justifying
beliefs. As preregistered, we modeled this relationship with two fixed
effects, one estimating the effect of system-justifying beliefs on
judgments about system-defending protest actions and one estimating
their effect on judgments about system-challenging protest actions, and
one varying effect estimating its variance across conditions. Supporting
\emph{Hypothesis 2b}, participants who \emph{rejected} system-justifying
beliefs were \emph{more} likely to consider system-challenging protest
actions (for defunding the police) to be acceptable means of protest
(\(\beta_{xy} = 0.38, [0.16, 0.58]\)). We did not, however, find
evidence for the ideological symmetry implied by \emph{Hypothesis 2b}
since participants who \emph{endorsed} system-justifying beliefs were
\emph{not} more likely to consider system-defending protest actions
(against defunding the police) to be acceptable means of protest
(\(\beta_{xy} = -0.18, [-0.40, 0.02]\)).

@@

We explored whether group identification moderated how the participants'
and the protesters' group memberships affected participants' responses.
To that end, we extended Model 1 by estimating participants' responses
as a function of their \emph{z}-standardized identification with their
racial ingroup. As in Model 2, we modeled this relationship with a fixed
and a varying effect. We found, however, that even participants who
strongly identified with their racial ingroup (+1\emph{SD}) did not
consider protest actions performed by their ingroup to be more
acceptable than the same actions performed by the relevant outgroup
(\(d = 0.01, [-0.00, 0.02]\)) and did not consider protest actions for a
cause that was aligned with their ingroup's interests to be more
acceptable than protest actions for a cause not aligned with their
ingroup's interests (\(d = 0.00, [-0.01, 0.01]\)). Our non-preregistered
analyses thus suggested that group identification did not moderate group
differences in judgments about collective action by ingroup and outgroup
members.

@@

In Experiment 3, we investigated potential double standards in judging
collective action for or against restricting abortion in the United
States. That is, we tested whether liberal and conservative women and
men applied different standards when judging collective action opposing
or supporting further restricting abortion. We conducted this study in
June 2022 when many expected the Supreme Court to overturn its \emph{Roe
v. Wade} decision that hitherto had guaranteed access to legal abortions
in the first 20--24 weeks of pregnancy in the United States. We focused
on restricting abortion as it is a conservative issue, as it primarily
affects women's reproductive rights, and as the anticipated Supreme
Court decision would prompt several states to restrict or ban abortion.
